Characterising Optimizations to Memory Access Patterns using Architecture-Independent Program Features

### 2020-01-19 paper/paper.tex
minor changes in wording for abstract and intro
Also changed 'ise' to 'ize' globally: see
https://theconversation.com/its-time-to-recognize-and-internalize-the-us-suffix-ize-19828

\raggedbottom
\acmConference[IWOCL'20]{International Workshop on OpenCL}{April 27--29, 2020}{Munich, Germany}
%\editor{John B. Appleseed}
\title[Architecture Independent Memory Access Pattern Analysis]{Characterising Optimisations to Memory Access Patterns using Architecture Independent Program Features}

@@

\keywords{Architecture Independent Analysis, Heterogenous Computing, Workload Characterisation, Memory Access Patterns} %TODO

@@

	%This paper addresses the challenge of profiling a code base to help programmers effectively optimise their codes for high performance.
	HPC developers are faced with the challenge of optimising OpenCL workloads for high performance on novel architectures.
	The Architecture Independent Workload Characterisation (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
	We propose the concept of parallel spatial locality -- the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
	We implement the parallel spatial locality metric in the AIWC framework, and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
	The differences in the obtained parallel spatial locality metric across implementations of matrix multiply reflect the optimisations performed.
	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalise to alternative or future systems.
	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimising codes for any given hardware target using the architecture independent workload characterisation (AIWC) tool -- specifically in profiling memory access patterns of code. 
	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimisations.
	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimisation of codes for arbitrary hardware targets.
	\end{abstract}

@@

	High-performance computing (HPC) systems are increasingly heterogeneous.
	A single node on a modern supercomputers may combine a traditional CPUs with one or more accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), or many integrated core devices (MICs).
	The development of high bandwidth interconnects between compute devices supports tight integration between multiple devices of different types on the same compute node.

	The OpenCL programming language is designed to support modern HPC software engineers in writing code that executes on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialised computational requirements aimed at FPGAs -- all in a single code base using OpenCL. 
	The language provides support for individual \emph{kernels}, portions of parallelisable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}. This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

	Application codes differ in resource requirements, control structure and available parallelism. Similarly, compute devices differ in number and capabilities of execution units, processing model, and available resources. These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimised systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by these kernels \cite{spafford2010maestro}.
	However this opportunity raises the twin challenges of determining the best hardware target for complex kernels and optimising codes to run on these diverse architectures.
	Our work is directed at helping developers tackle the latter of these two problems.

	Modern computer architectures have suffered from the exponential growth in computational speeds being unmet by proportional increases in memory speeds\cite{hennessycomparch}. Memory accesses pose larger bottlenecks to performance as applications' demand for main memory scales with the arithmetic capability of computer systems. To mitigate this latency, modern CPU designs have employed a wide range of cache technologies to reduce main memory accesses using the principle of \emph{spatial locality}: the idea that programs repeatedly access memory addresses in close locality. On the other hand, GPUs rely on hardware multithreading to hide memory latency, and their architectures favour ALU capability over sophisticated logic to manage a cache hierarchy and out-of-order execution. As a result, the performance of kernels on CPUs and GPUs alike is strongly dependent on memory access patterns intrinsic to the code. % We discuss the memory model design and the manufacturer recommended optimisations for NVIDIA GPUs in later sections..

	%We demonstrate the effectiveness of an architecture-independent analysis of codes in determining ... Our work guides optimisation efforts for OpenCL codes running on increasingly diverse hardware by examining the current state of the art in architecture-independent metrics for memory access patterns.

	%What we did briefly
	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware dependent code optimisation -- specifically by guiding the improvement of codes' memory access patterns. We provide examples of manufacturer recommended code optimisations to improve memory access patterns on the target architecture, and examine these using AIWC \cite{beauaiwc}, a state of the art architecture-independent code analysis tool for parallel workloads. Our work highlights the benefits and challenges arising from an architecture independent analysis of memory based program characteristics. We suggest the addition of two new metrics to AIWC to better characterise these memory-based optimisations. We measure the presented codes using the new metrics and discuss the metrics' effectiveness in capturing the essence of the optimisations performed. With these improvements, we show that AIWC is better suited to be used by HPC software engineers in guiding optimisation of their codes.

	The structure of this paper is as follows. In section \ref{motivating example}, we use the example of GPU matrix multiplication code to showcase the specific vendor recommended optimisations our work aims to measure. In section \ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture independent fashion, and relevant architecture dependent approaches to memory access profiling. Then in section \ref{method} we provide a methodology to critically examine AIWC, the current state of the art architecture independent workload characterisation tool. In section \ref{results} we measure the accuracy of AIWC in profiling the impact of performance optimisations to various OpenCL codes. In section \ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware specific memory optimisations. Finally, in section \ref{future work} we discuss further avenues to extend our work and conclude.

@@

	We aim to capture the essence of vendor recommended optimisations for target architectures in our metrics. We demonstrate the optimisation strategies being analysed by presenting a simple unoptimised matrix multiplication kernel for input square matrices of order $N$, and then improve the memory access patterns of this kernel by incrementally performing optimisations recommended in the CUDA Optimisation Handbook \cite{cudaoptimisation} for NVIDIA GPUs. These incrementally optimised OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.

	\subsection{Simple unoptimised matrix multiplication}

	The unoptimised matrix multiplication kernel presented in Appendix \ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimisation. Each thread of the kernel computes the $(globalRow, globalCol)$ element of matrix $C$ by reading the appropriate row and column from matrices $A$ and $B$ respectively.

@@

	The code is further optimised by improving how the previous \texttt{coalescedAMultiply} kernel handles accesses to Matrix $B$. While each thread in coalescedAMultiply accesses only a single element of matrix $A$ from global memory, each thread accesses a full column of matrix $B$. The repeated reads to elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory. Appendix \ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimisation.

@@

	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel accesses to shared memory. Bank conflicts occur when shared memory in the same bank is accessed concurrently. The code in \texttt{coalescedAB} kernel is further optimised by implicitly transposing tiles of Matrix $A$ while loading from global memory. This improves memory bank utilisation during reads to shared memory tiles are performed during the necessary computation. The \texttt{coalescedABT} kernel demonstrates this optimisation by modifying lines 20,21 of \texttt{coalescedAB} to:

@@

	Memory access alignment is important to best utilise all parts of GPU memory architecture. Global memory buffer alignment can allow threads to access blocks of global memory aligned to the nearest cache line. This enables coalescing of memory accesses. If buffers are misaligned, parallel memory requests may cross over cache lines, and may double the number of slow global memory accesses needed as demonstrated in figures \ref{fig: Simple memory alignment} and \ref{fig: Simple memory misalignment}.

@@

	All the examples above are an incomplete listing of optimisation strategies developers can perform to code targeted at NVIDIA GPUs. We analyse the correlations between the collected AIWC metrics for each of these kernels and our conceptual understanding of GPU memory optimisations. This allows us to examine how effectively our methodology uncovers underlying bottlenecks and guides optimisation efforts.

@@

	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks. A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimise these codes. Architecture dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates. These are typically collected using hardware performance counters available on most target architectures. These performance counters do not serve to guide optimisation beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}. Further, in many cases, architecture dependent characteristics usually cannot be directly correlated to specific code patterns. For example, the causes of high cache miss rates in the execution of a program are too complex and widely variant on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies. In summary, we argue that a HPC developer tasked with optimising code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimisation patterns. Such architecture-independent metrics will effectively guide the HPC developer's workflow in both finding and fixing bottlenecks in code performance.

	The Architecture Independent Workload Characterisation (AIWC) tool \cite{beauaiwc} relies on a set of instruction set architecture (ISA)-independent features determined by Shao and Brooks \cite{shao2013isa}. AIWC measures memory based metrics to characterise the target code's memory behaviours. AIWC firstly collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses. A small ratio of 90\% memory footprint to total memory footprint correlates to programs accessing the same small subset of memory addresses repeatedly -- which can be linked to increased performance in a cached memory hierarchy.

@@

	LMAE measures the locality of memory accesses performed over the full execution of a program. A steeper drop in entropy with increasing number of bits may correlate to more localised memory accesses over the program's execution.

@@

	\subsection{Advancements in Memory Access Pattern Characterisation}

	We provide a brief summary of past work in architecture independent program characterisation and Memory Access Pattern analysis:

	\begin{itemize}
		\item The Microarchitecture Independent Workload Characterisation Tool (MICA) \cite{hoste2007microarchitecture} was the first microarchitecture independent workload characterisation tool for single threaded programs. 
		\item The Workload ISA-Independent Characterization for Applications (WIICA) \cite{shao2013isa} extends MICA to present a framework to analyse single threaded programs independent of the ISA.
		\item CuMAPz: CuMAPz \cite{kim2011cumapz} is a CUDA memory access profiling tool and the first to present metrics for CUDA codes which can be generalised to any given NVIDIA GPU.
	\end{itemize}

@@

	The MICA framework was developed by Hoste and Eeckout \cite{hoste2007microarchitecture} in response to their findings on the limitations of conventional microarchitecture dependent characteristics. Hoste and Eeckout discovered that performance counter based approaches to profiling codes often struggled in finding underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture. The MICA framework is a holistic characterisation tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.

	Of these metrics, data stream strides are of particular interest in memory access pattern profiling. MICA's stride length metric measures the distance between consecutive memory accesses in a single threaded application. For CPU architectures running single threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times. This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}. %Intel optimisation manual.

@@

	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access pattern analysis tool to guide NVIDIA GPU optimisations. CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimisations. CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.

@@

	Our approach is an architecture independent analysis of memory access patterns that provides metrics correlating to similar performance critical memory access optimisations as CuMAPz. We aim to further the state of the art by providing a framework to guide developers in optimising OpenCL codes for any given target architecture. To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterise memory access patterns of parallel applications independent of the target architecture.

@@

	To provide memory access pattern metrics that represent vendor recommended optimisations on GPUs, we improved AIWC by adding a new metric: relative local memory usage. This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}. On GPUs, this memory address space is mapped to fast on-chip shared memory.

@@

	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimisations to the matrix multiplication code lead to increased performance. Note that a matrix size of 80 is small -- overheads attached to optimisation techniques may be causing the observed decrease in performance as some optimisations are performed.

@@

	Table \ref{matmul aiwc table} summarises AIWC memory metrics collected for each of the matrix multiply kernels in section \ref{motivating example}. Note that only two LMAE values are shown for brevity. We now analyse the effectiveness of AIWC metrics in profiling the NVIDIA recommended memory optimisations applied to the matrix multiplication kernel.

@@

	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}. This is explained by the optimisations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in on-chip shared memory. We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices. The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localised shared memory buffers. The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small shared memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.

	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simpleMultiply} to $0.25\%$ (\texttt{coalescedAB}). Increased utilisation of shared memory in the optimised kernels means that the shared memory buffers make up a greater proportion of total memory accesses in the program. As the shared memory buffers are small and reused within a workgroup, the memory footprint of the shared memory buffers is also smaller.

	AIWC's metrics strongly reward optimisations that tend to localise memory accesses. Shared memory buffers are typically smaller than global memory arrays when programming for GPUs due to hardware limitations on sizes of shared memory buffers \cite{cudamanual}. However, the metrics currently measured by AIWC do not have a direct causal relation to code patterns that optimise memory accesses on GPUs. The proposed relative local memory usage metric is the first to correspond to a recommended optimisation strategy of using fast on-chip shared memory. Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB}, \texttt{coalescedABT} and \texttt{alignedABT} codes. We address the root cause of this -- a lack of temporal information about memory accesses, in the next section.%TODO READ THIS SECTION

@@

	We propose a new architecture independent metric to measure memory access patterns of parallel programs. The proposed metric is inspired by CuMAPz' direct approaches to measuring optimisation specific characteristics of CUDA codes.

@@

	\textit{Summarise}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group. We then average the n-bits dropped entropy summaries across thread groups to obtain the \textit{n-bits dropped parallel spatial locality metrics} for the kernel's execution.

@@

	In figure \ref{fig: matmul parallel spatial locality} we see theoretical calculations for the proposed parallel spatial locality metric for each of the matrix multiplication kernels from section \ref{motivating example}. In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses. It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localised simply due to the small size of shared on-chip memory typically available on GPUs. Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels. It accurately depicts a steeper drop-off for the more optimised \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels. This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localise shared memory accesses to better use the hardware provided.

@@

	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorise 
	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.

>>>>

\raggedbottom
\acmConference[IWOCL'20]{International Workshop on OpenCL}{April 27--29, 2020}{Munich, Germany}
%\editor{John B. Appleseed}
\title[Architecture-Independent Memory Access Pattern Analysis]{Characterising Optimizations to Memory Access Patterns using Architecture-Independent Program Features}

@@

\keywords{Architecture Independent Analysis, Heterogenous Computing, Workload Characterization, Memory Access Patterns} %TODO

@@

	%This paper addresses the challenge of profiling a code base to help programmers effectively optimize their codes for high performance.
	HPC developers are faced with the challenge of optimizing OpenCL workloads for high performance on novel architectures.
	The Architecture-Independent Workload Characterization (AIWC) tool is a plugin for the Oclgrind OpenCL simulator that gathers metrics of OpenCL programs used to understand and predict program performance on an arbitrary given hardware architecture.
	However, AIWC metrics are not always easily interpreted and do not reflect some important memory access patterns affecting efficiency across architectures.
	We propose the concept of parallel spatial locality: the locality of memory accesses by OpenCL work-items (threads) executing simultaneously.
	We implement the parallel spatial locality metric in the AIWC framework and analyse gathered results on matrix multiply and the Extended OpenDwarfs OpenCL benchmarks.
	The differences in the parallel spatial locality metric between implementations of matrix multiply reflect the optimizations performed.
	The new metric showcases the memory access patterns in the OpenDwarfs benchmarks that affect their performance on various architectures.
	%Current methods of analysing program behaviour have largely relied on performance counters on various architectures to provide architecture dependent metrics of code behaviour. 
	%However, if workload characteristics are tied to architectural features specific to a particular system, these analysis frameworks may not generalize to alternative or future systems.
	%Considering the trend towards heterogeneous computing, we work towards a framework to guide developers in optimizing codes for any given hardware target using the architecture independent workload characterization (AIWC) tool -- specifically in profiling memory access patterns of code. 
	%Metrics must be easily interpreted and conceptually linked to patterns in hardware-specific memory access optimizations.
	%To this end we propose, implement and analyse the addition of two new metrics to AIWC and present a new approach to profile spatial locality intrinsic to parallel codes.
	The improvements suggested to AIWC will help HPC developers better understand memory access patterns of complex codes, a crucial step in guiding optimization of codes for arbitrary hardware targets.
	\end{abstract}

@@

	High-performance computing (HPC) systems are increasingly heterogeneous.
	A single node on a modern supercomputer may combine traditional CPUs with one or more accelerators such as GPUs, Field Programmable Gate Arrays (FPGAs), or many integrated core devices (MICs).
	High bandwidth interconnects support tight integration between multiple devices of different types on a single compute node.

	The OpenCL programming language is designed to support portability for heterogeneous systems by allowing a single code to execute on multiple hardware targets. 
	%It is possible for an engineer to write highly thread-divergent logical analysis code aimed at traditional CPUs; parallelisable data crunching kernels aimed at GPUs; and code to satisfy specialized computational requirements aimed at FPGAs -- all in a single code base using OpenCL. 
	The language provides support for individual \emph{kernels}, portions of parallelizable codes that make up a HPC code base, to be executed independently on diverse hardware targets such as traditional CPUs, GPUs and FPGAs \cite{khronosopenclworkinggroup2011opencl}. This gives HPC software developers greater flexibility by allowing codes aimed at a range of hardware targets to be written in a single programming language environment.

	Application codes differ in resource requirements, control structure and available parallelism. Similarly, compute devices differ in number and capabilities of execution units, processing model, and available resources.
	These heterogeneous computing environments present opportunities for software engineers and HPC integrators to design highly optimized systems with multiple kernels executing on hardware targets best suited for the diverse computational tasks performed by each kernel~\cite{spafford2010maestro}.
	However, this opportunity also presents a challenge in optimizing code to run on diverse architectures.
	The work reported here aims to help HPC developers understand and optimize memory access patterns to improve performance on heterogeneous systems.

	In the evolution of computer architectures over the last few decades, the exponential growth in computational capability has not been matched by proportional increases in memory speeds\cite{hennessycomparch}.
	Memory accesses pose larger bottlenecks to performance as application demand for main memory scales with the arithmetic capability of computer systems.
	To mitigate this latency, modern CPU designs have employed a wide range of cache technologies to reduce main memory accesses using the principle of \emph{spatial locality}: the observation that data that a program accesses close together in time tend to also be close together in memory.
	On the other hand, GPUs rely on hardware multithreading to hide memory latency, and their architectures favour ALU capability over sophisticated logic to manage a cache hierarchy and out-of-order execution. As a result, the performance of kernels on CPUs and GPUs alike is strongly dependent on memory access patterns intrinsic to the code. % We discuss the memory model design and the manufacturer recommended optimizations for NVIDIA GPUs in later sections..

	%We demonstrate the effectiveness of an architecture-independent analysis of codes in determining ... Our work guides optimization efforts for OpenCL codes running on increasingly diverse hardware by examining the current state of the art in architecture-independent metrics for memory access patterns.

	%What we did briefly
	Our aim in this paper is to develop a framework to guide HPC software engineers in hardware-dependent code optimization -- specifically by guiding the improvement of memory access patterns.
	We provide examples of manufacturer-recommended code optimizations to improve memory access patterns on the target architecture, and examine these using the architecture-independent workload characterization (AIWC)~\cite{beauaiwc} plugin for Oclgrind.
	Our work highlights the benefits and challenges arising from an architecture-independent analysis of memory-based program characteristics.
	We propose two new metrics to AIWC to better characterize these memory-based optimizations.
	We measure the presented codes using the new metrics and demonstrate the metrics' effectiveness in capturing the essence of the optimizations performed.

	The structure of this paper is as follows.
	In section~\ref{motivating example} we use the example of matrix multiplication on a GPU to showcase the specific vendor-recommended optimizations our work aims to measure.
	In section~\ref{related work} we discuss how AIWC and its precursors profiled memory access patterns in an architecture-independent fashion, and relevant architecture dependent approaches to memory access profiling.
	In section \ref{method} we present a methodology to critically examine the architecture-independent workload characterization tool, which we use in section~\ref{results} to measure the accuracy of AIWC in profiling the impact of performance optimizations to various OpenCL codes.
	In section~\ref{discussion} we suggest new methods to accurately capture code patterns correlating to hardware-specific memory optimizations.
	Finally, in section~\ref{future work} we discuss further avenues to extend our work and conclude.

@@

	We aim to capture the essence of vendor recommended optimizations for target architectures in our metrics.
	We demonstrate the optimization strategies being analysed by presenting a simple unoptimized matrix multiplication kernel for input square matrices of order $N$, and then improve the memory access patterns of this kernel by incrementally performing optimizations recommended in the CUDA Optimization Handbook \cite{cudaoptimisation} for NVIDIA GPUs.
	These incrementally optimized OpenCL codes are then analysed using AIWC to determine the accuracy in profiling favourable memory access patterns of the current framework and the proposed extensions to AIWC.

	\subsection{Simple unoptimized matrix multiplication}

	The unoptimized matrix multiplication kernel presented in Appendix \ref{simpleMultiply} is used as a baseline to validate the performance improvements measured from each following optimization. Each thread of the kernel computes the $(globalRow, globalCol)$ element of matrix $C$ by reading the appropriate row and column from matrices $A$ and $B$ respectively.

@@

	The code is further optimized by improving how the previous \texttt{coalescedAMultiply} kernel handles accesses to Matrix $B$. While each thread in coalescedAMultiply accesses only a single element of matrix $A$ from global memory, each thread accesses a full column of matrix $B$. The repeated reads to elements of matrix $B$ across multiple threads can be alleviated by reading tiles from matrix $B$ into shared memory. Appendix \ref{coalescedABMultiply} contains the \texttt{coalescedAB} kernel which performs this optimization.

@@

	NVIDIA shared memory is divided into multiple banks -- stored in independent memory modules -- to allow parallel accesses to shared memory. Bank conflicts occur when shared memory in the same bank is accessed concurrently. The code in \texttt{coalescedAB} kernel is further optimized by implicitly transposing tiles of Matrix $A$ while loading from global memory. This improves memory bank utilization during reads to shared memory tiles are performed during the necessary computation. The \texttt{coalescedABT} kernel demonstrates this optimization by modifying lines 20,21 of \texttt{coalescedAB} to:

@@

	Memory access alignment is important to best utilize all parts of GPU memory architecture. Global memory buffer alignment can allow threads to access blocks of global memory aligned to the nearest cache line. This enables coalescing of memory accesses. If buffers are misaligned, parallel memory requests may cross over cache lines, and may double the number of slow global memory accesses needed as demonstrated in figures \ref{fig: Simple memory alignment} and \ref{fig: Simple memory misalignment}.

@@

	All the examples above are an incomplete listing of optimization strategies developers can perform to code targeted at NVIDIA GPUs. We analyse the correlations between the collected AIWC metrics for each of these kernels and our conceptual understanding of GPU memory optimizations. This allows us to examine how effectively our methodology uncovers underlying bottlenecks and guides optimization efforts.

@@

	Hoste and Eeckout \cite{hoste2007microarchitecture} show that while conventional architecture dependent characteristics are useful in locating performance bottlenecks, they can hide the underlying, inherent program behaviour causing performance bottlenecks. A conceptual understanding of the performance characteristics of complex codes is necessary for the programmer to effectively optimize these codes. Architecture dependent characteristics typically include instructions per cycle (IPC), cache and branch prediction miss rates, page faults and DRAM bus data transfer rates. These are typically collected using hardware performance counters available on most target architectures. These performance counters do not serve to guide optimization beyond highlighting potential bottlenecks \cite{hoste2007microarchitecture, ganesan2008performance}. Further, in many cases, architecture dependent characteristics usually cannot be directly correlated to specific code patterns. For example, the causes of high cache miss rates in the execution of a program are too complex and widely variant on microarchitecture specific features such as cache size, prefetch behaviour and cache placement policies. In summary, we argue that a HPC developer tasked with optimizing code for a given hardware target would benefit from architecture-independent metrics of the code that can be conceptually correlated to hardware specific optimization patterns. Such architecture-independent metrics will effectively guide the HPC developer's workflow in both finding and fixing bottlenecks in code performance.

	The Architecture-Independent Workload Characterization (AIWC) tool \cite{beauaiwc} relies on a set of instruction set architecture (ISA)-independent features determined by Shao and Brooks \cite{shao2013isa}. AIWC measures memory based metrics to characterize the target code's memory behaviours. AIWC firstly collects aggregate simple counts such as the (i) total memory footprint, the total number of unique addresses accessed; and (ii) 90\% memory footprint, the number of unique addresses covering 90\% of memory accesses. A small ratio of 90\% memory footprint to total memory footprint correlates to programs accessing the same small subset of memory addresses repeatedly -- which can be linked to increased performance in a cached memory hierarchy.

@@

	LMAE measures the locality of memory accesses performed over the full execution of a program. A steeper drop in entropy with increasing number of bits may correlate to more localized memory accesses over the program's execution.

@@

	\subsection{Advancements in Memory Access Pattern Characterization}

	We provide a brief summary of past work in architecture independent program characterization and Memory Access Pattern analysis:

	\begin{itemize}
		\item The Microarchitecture Independent Workload Characterization Tool (MICA) \cite{hoste2007microarchitecture} was the first microarchitecture independent workload characterization tool for single threaded programs. 
		\item The Workload ISA-Independent Characterization for Applications (WIICA) \cite{shao2013isa} extends MICA to present a framework to analyse single threaded programs independent of the ISA.
		\item CuMAPz: CuMAPz \cite{kim2011cumapz} is a CUDA memory access profiling tool and the first to present metrics for CUDA codes which can be generalized to any given NVIDIA GPU.
	\end{itemize}

@@

	The MICA framework was developed by Hoste and Eeckout \cite{hoste2007microarchitecture} in response to their findings on the limitations of conventional microarchitecture dependent characteristics. Hoste and Eeckout discovered that performance counter based approaches to profiling codes often struggled in finding underlying program features that map to improved or worsened usage of performance critical hardware features of the target architecture. The MICA framework is a holistic characterization tool, and thus collects features including instruction mix, instruction-level parallelism, register traffic, data stream strides and branch predictability.

	Of these metrics, data stream strides are of particular interest in memory access pattern profiling. MICA's stride length metric measures the distance between consecutive memory accesses in a single threaded application. For CPU architectures running single threaded applications, this metric correlates to the spatial locality of memory accesses -- a measure of how closely bunched are memory access in nearby times. This is directly correlated to cache reuse rates, critical to code performance on CPUs \cite{inteloptimisation}. %Intel optimization manual.

@@

	Kim and Shrivastava \cite{kim2011cumapz} present CuMAPz, a CUDA memory access pattern analysis tool to guide NVIDIA GPU optimizations. CuMAPz focuses on the problem of improving CUDA application performance using NVIDIA memory-based optimizations. CuMAPz analyses CUDA codes structurally and simulates code execution on the memory hierarchy of specific NVIDIA GPU models.

@@

	Our approach is an architecture independent analysis of memory access patterns that provides metrics correlating to similar performance critical memory access optimizations as CuMAPz. We aim to further the state of the art by providing a framework to guide developers in optimizing OpenCL codes for any given target architecture. To the best our knowledge, none of the previous works present a set of performance metrics that accurately characterize memory access patterns of parallel applications independent of the target architecture.

@@

	To provide memory access pattern metrics that represent vendor recommended optimizations on GPUs, we improved AIWC by adding a new metric: relative local memory usage. This measures the proportion of all memory accesses from the symbolic execution of the kernel that occurred to memory allocated as \texttt{\_\_local}. On GPUs, this memory address space is mapped to fast on-chip shared memory.

@@

	Figure \ref{fig: matmul runtime} shows that across varying problem sizes, the vendor recommended optimizations to the matrix multiplication code lead to increased performance. Note that a matrix size of 80 is small -- overheads attached to optimization techniques may be causing the observed decrease in performance as some optimizations are performed.

@@

	Table \ref{matmul aiwc table} summarizes AIWC memory metrics collected for each of the matrix multiply kernels in section \ref{motivating example}. Note that only two LMAE values are shown for brevity. We now analyse the effectiveness of AIWC metrics in profiling the NVIDIA recommended memory optimizations applied to the matrix multiplication kernel.

@@

	\textit{Global and local MAE}: Entropy measurements decrease from \texttt{simpleMultiply} to \texttt{coalescedAB}. This is explained by the optimizations performed incrementally reducing accesses to matrix $A$ and $B$, and instead performing the bulk of computations on smaller tiles in on-chip shared memory. We observe an almost completely uniform distribution of memory accesses in \texttt{simpleMultiply}, where the program makes $N$ loads to each element of $A$ and $B$ with $N$ the dimension of the matrices. The frequency distribution of memory addresses accessed become increasingly non-uniform as we perform less accesses to matrices $A$ and $B$ and offset this by increasing the frequency of accesses to smaller, more localized shared memory buffers. The non-uniformity of relative frequency with which each memory cell is accessed by the program translates directly to decreases in local and global memory entropy values as reliance on small shared memory buffers increases from \texttt{simpleMultiply} to \texttt{coalescedAB}.

	\textit{Memory Footprint}: Similar to trends in global and local MAE, we find that the ratio of 90\% memory footprint to total memory footprint decreases from $60.12\%$ for \texttt{simpleMultiply} to $0.25\%$ (\texttt{coalescedAB}). Increased utilization of shared memory in the optimized kernels means that the shared memory buffers make up a greater proportion of total memory accesses in the program. As the shared memory buffers are small and reused within a workgroup, the memory footprint of the shared memory buffers is also smaller.

	AIWC's metrics strongly reward optimizations that tend to localize memory accesses. Shared memory buffers are typically smaller than global memory arrays when programming for GPUs due to hardware limitations on sizes of shared memory buffers \cite{cudamanual}. However, the metrics currently measured by AIWC do not have a direct causal relation to code patterns that optimize memory accesses on GPUs. The proposed relative local memory usage metric is the first to correspond to a recommended optimization strategy of using fast on-chip shared memory. Further, we find that all current AIWC metrics do not measure any sizeable difference between \texttt{coalescedAB}, \texttt{coalescedABT} and \texttt{alignedABT} codes. We address the root cause of this -- a lack of temporal information about memory accesses, in the next section.%TODO READ THIS SECTION

@@

	We propose a new architecture independent metric to measure memory access patterns of parallel programs. The proposed metric is inspired by CuMAPz' direct approaches to measuring optimization specific characteristics of CUDA codes.

@@

	\textit{Summarize}: average the collected entropy values across all the timestamps to calculate the parallel spatial locality metrics for one thread group. We then average the n-bits dropped entropy summaries across thread groups to obtain the \textit{n-bits dropped parallel spatial locality metrics} for the kernel's execution.

@@

	In figure \ref{fig: matmul parallel spatial locality} we see theoretical calculations for the proposed parallel spatial locality metric for each of the matrix multiplication kernels from section \ref{motivating example}. In particular we observe that the \texttt{coalescedABT} and \texttt{alignedABT} kernels have the steepest drop-offs in entropy as the number of bits skipped is increased, which correlates to better locality of parallel memory accesses. It is expected for these kernels to exhibit better parallel spatial locality than \texttt{simpleMultiply}, as \texttt{coalescedABT} and \texttt{alignedABT} make use of shared memory, where accesses tend to be localized simply due to the small size of shared on-chip memory typically available on GPUs. Further we find that the proposed metric successfully distinguishes between the \texttt{coalescedAB} and \texttt{coalescedABT} kernels. It accurately depicts a steeper drop-off for the more optimized \texttt{coalescedABT} kernel, where a larger proportion of parallel memory accesses make better use of the memory bank structure of GPU shared memory than all previous kernels. This is a significant improvement over the state-of-the-art AIWC metrics in characterising how codes localize shared memory accesses to better use the hardware provided.

@@

	The Extended OpenDwarfs benchmark suite is a set of OpenCL workload classified according to the computation patterns or Dwarfs. First presented in 2006, the Dwarfs categorize 
	We present the results of running the AIWC tool on the Extended OpenDwarfs benchmark suite \cite{opendwarfs2017base,extendedopendwarfs}.
